---
title: "GRASS-GIS tutorial"
author: "Lara Katz"
date: "2024-05-07"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Libraries
```{r message=FALSE}
# install.packages("rgrass") # rgrass7 in tutorial but rgrass is the new version
# install.packages("link2GI")
library(rgrass)
library(link2GI)
library(tmap)
library(dplyr)
library(terra)
library(sf)
```

# Set up GRASS workspace

```{r}
filepath <- "D:/GRASS GIS 8.3/"
projectpath <- "D:/GRASS GIS 8.3/NortheastDEM/"
rgrass::initGRASS(gisBase = filepath, gisDbase = projectpath, mapset = "PERMANENT")
stringexecGRASS("g.proj -c --quiet epsg=26919") # set projection
```

### Notes on GRASS:

**Sources:** 
* https://florisvdh.github.io/presentations/20210617_grass/

* https://baharmon.github.io/watersheds-in-grass

* https://www.youtube.com/watch?v=y4mz_nt1TaU

**Wrapper functions:**
Run GRASS commands in R:
`execshell("g.proj -c --quiet epsg=31370")`
`stringexecGRASS("g.proj -c --quiet epsg=31370")` 

Breaking up command as arguments of execGRASS():
`execGRASS("g.proj", flags = c("c", "quiet"), epsg = 31370)`
`execshell()`: advantage of accepting entire GRASS script; supports maximum number of commands
`rgrass7::execGRASS()`: advantage that the GRASS command parameters can be more easily programmed in R
`rgrass7::stringexecGRASS()`: advantage of using original GRASS syntax (e.g. copied from GRASS session), while following rgrass7 philosophy of handling user input in R

Set to using either `sp` or `sf+stars` with `use_sp()` or `use_sf()` (although this doesn't seem to work with the new package version).

# Load and combine DEM rasters

Download and import DEM rasters from TNM:
```{r}
# files <- list.files(path = projectpath,
#                     pattern = "HUC6_10m",
#                     all.files = TRUE,
#                     full.names = TRUE,
#                     include.dirs = FALSE)
# 
# for (i in 1:length(files)){
#   stringexecGRASS(paste0(paste0(paste0("r.import resample=bilinear input='",files[i]), "' output=dem"), i))
# }
stringexecGRASS(paste0(paste0("r.import resample=bilinear input='",projectpath), "Elevation_HUC6_30m.tif' output=Elevation_HUC6_30m")) # no need to reproject, already in UTM 19N
```


Patch DEM rasters together:
```{r}
# rasters <- "dem1@PERMANENT"
# for (i in 2:length(files)){
#   rasters <- paste(rasters, paste0(paste0("dem",i), "@PERMANENT"), sep = ",")
# }
# rasters
# 
# stringexecGRASS(paste0(paste0("r.patch --overwrite input=", rasters), " output=DEM_all@PERMANENT"))
```

Export raster into R:
```{r}
# stringexecGRASS(paste0(paste0("r.out.gdal input=DEM_all@PERMANENT output=", projectpath), "DEM_all.tif"))
```

# Extent shapefile

Load shapefile of 200km buffer around Saco and Merrimack HUC6
```{r}
range <- vect("./Inputs_raw/SacoMerrimackHUC6.shp")
range <- project(range, y = "epsg:26919") # EPSG for NAD83 / UTM Zone 19N
range <- aggregate(range, dissolve = TRUE) # dissolve into 1 feature
range.buff50 <- terra::buffer(range, 50000)
writeVector(range.buff50,paste0(projectpath, "SacoMerrimack50kmbuff.shp"))
range.buff100 <- terra::buffer(range, 100000) # buffer to 200,000m or 200km
writeVector(range.buff100,paste0(projectpath, "SacoMerrimack100kmbuff.shp"))

extent <- st_read(paste0(projectpath, "SacoMerrimack100kmbuff.shp"))
# canada <- st_read(paste0(projectpath, "Canada_mask_bbox.shp"))
```


Use HUC10 polygons from Maine, New Hampshire, and New York within WBD01 and a 100km buffer of the Saco and Merrimack HUC6s (GRASS can't compute anything larger than this)
```{r}
# Maine, NH, MA
hucs_east1 <- st_read("WBD_01_HU2_GDB.gdb", layer="WBDHU10") %>% st_transform(.,crs=st_crs(extent))
# hucs_me2 <- st_read("NHD_H_Maine_State_GDB.gdb", layer="WBDHU10") %>% st_transform(.,crs=st_crs(extent))
hucs_east <- hucs_east1[extent,] %>% st_union(.) %>% st_as_sf(.)
# hucs_me <- hucs_me[hucs_me$name != "Chaudiere River",] # only a tiny part of HUC in 
plot(hucs_east[1])

# # NY, VT, MA
# hucs_south1 <- st_read("./Shape/WBDHU10.shp") %>% st_transform(.,crs=st_crs(extent))
# # hucs_me2 <- st_read("NHD_H_Maine_State_GDB.gdb", layer="WBDHU10") %>% st_transform(.,crs=st_crs(extent))
# hucs_south <- hucs_south1[extent,] %>% st_union(.) %>% st_as_sf(.)
# # hucs_me <- hucs_me[hucs_me$name != "Chaudiere River",] # only a tiny part of HUC in 
# plot(hucs_south[1])
# # Massachusetts
# # hucs_ma1 <- st_read("NHD_H_Massachusetts_State_GDB.gdb", layer="WBDHU10") %>% st_transform(.,crs=st_crs(extent))
# # hucs_ma <- hucs_ma1[extent,] # within buffer
# #   # st_union(.) %>% st_as_sf(.)
# 
# hucs_west1 <- st_read("./WBD_04_HU2_Shape/Shape/WBDHU10.shp") %>% st_transform(.,crs=st_crs(extent))
# # hucs_me2 <- st_read("NHD_H_Maine_State_GDB.gdb", layer="WBDHU10") %>% st_transform(.,crs=st_crs(extent))
# hucs_west <- hucs_west1[extent,] %>% st_union(.) %>% st_as_sf(.)
# # hucs_me <- hucs_me[hucs_me$name != "Chaudiere River",] # only a tiny part of HUC in 
# plot(hucs_west[1])


# # New Hampshire
# hucs_nh1 <- st_read("NHD_H_New_Hampshire_State_GDB.gdb", layer="WBDHU10") %>% st_transform(.,crs=st_crs(extent))
# hucs_nh <- hucs_nh1[extent,] # within buffer
#   # st_union(.) %>% st_as_sf(.)
#   
# # New York
# hucs_ny1 <- st_read("NHD_H_New_York_State_GDB.gdb", layer="WBDHU10") %>% st_transform(.,crs=st_crs(extent))
# hucs_ny <- hucs_ny1[extent,] # within buffer
# # hucs_ny <- hucs_ny %>% filter(name != "Lower Hudson") 
#   # st_union(.) %>% st_as_sf(.)
# 
# # Vermont
# hucs_vt1 <- st_read("NHD_H_Vermont_State_GDB.gdb", layer="WBDHU10") %>% st_transform(.,crs=st_crs(extent))
# hucs_vt <- hucs_vt1[extent,] # within buffer
# hucs_ny <- hucs_ny %>% filter(name != "Lower Hudson") 
  # st_union(.) %>% st_as_sf(.)

hucs <- dplyr::bind_rows(list(hucs_south,hucs_west,hucs_east)) %>% st_union(.) %>% st_as_sf(.) %>% vect(.) %>% fillHoles(.)

hucs <- hucs_east %>% vect(.) %>% fillHoles(.)
plot(hucs)
#   dplyr::bind_rows(list(hucs_me,hucs_nh,hucs_ny)) %>% st_union(.) %>% st_as_sf(.)
# plot(hucs)
# hucs1 <- hucs %>% filter(name == "Saco" | name == "Merrimack" | name == "Connecticut" | name == "Androscoggin" | name == "Connecticut Coastal" | name == "Massachusetts-Rhode Island Coastal" | name == "Maine Coastal" | name == "Kennebec") %>% st_union(.) %>% st_as_sf(.) %>% vect(.) %>% fillHoles(.)
# 
# 
# hucs2 <- hucs %>% filter(name == "Penobscot" | name == "Lower Saint Lawrence River" | name == "Middle Saint Lawrence River" | name == "Upper Saint Lawrence River" | name == "Upper Hudson" | name == "Lower Hudson-Long Island") %>% st_union(.) %>% st_as_sf(.) %>% vect(.) %>% fillHoles(.)
# 
# plot(hucs1)
# plot(hucs2)

bbox <- st_as_sfc(st_bbox(hucs))
plot(bbox)
st_write(bbox,paste0(projectpath, "bbox.shp"), append=F)
```

Import extent shapefile:
```{r}
stringexecGRASS(paste0(paste0("v.import --overwrite input='", projectpath),"bbox.shp' layer=bbox output=BoundingBox"))
```

Set computational region:
```{r}
stringexecGRASS("g.region vector=BoundingBox@PERMANENT")
```


Clip top of HUCs along Canadian border since the elevation data is all NA there
```{r}
# hucs.mask <- vect(hucs)
# # canada.v <- vect(canada)
# # 
# # hucs.clip <- erase(hucs.v, canada.v)
# # 
# # plot(hucs.clip)
# writeVector(hucs.mask,paste0(projectpath, "HUC4.shp"), overwrite=T)
# 
# hucs.mask <- vect(paste0(projectpath, "HUC4.shp"))
```

Load DEM:
```{r}
dem <- rast(paste0(projectpath, "Elevation_HUC6_30m.tif"))
nrow <- nrow(dem)
ncol <- ncol(dem) 
```


Make a mask raster for each HUC4
```{r}
# for (i in 1:nrow(hucs.mask)){
#   r <- rast(hucs.mask[i],nrows=nrow, ncols=ncol) # empty raster the size of the DEM raster
#   z <- rasterize(hucs.mask[i], r, "name")
#   align <- resample(z, dem)
#   crop <- crop(align, ext(hucs.mask[i]))
#   writeRaster(crop, paste0(paste0(paste0(projectpath, "HUC4-"), hucs.mask[i]$huc4), ".tif"), overwrite=T)
# }

r <- rast(hucs,nrows=nrow,ncols=ncol)
z <- rasterize(hucs, r)
align <- resample(z, dem)
crop <- crop(align, ext(hucs))
writeRaster(crop, paste0(projectpath, "HUC10-100km.tif"), overwrite=T)
plot(crop)
# r <- rast(hucs2,nrows=nrow,ncols=ncol)
# z <- rasterize(hucs2, r)
# align <- resample(z, dem)
# crop <- crop(align, ext(hucs2))
# writeRaster(crop, paste0(projectpath, "HUC4-OuterHUCs.tif"), overwrite=T)
# test <- rast("D:/GRASS GIS 8.3/NortheastDEM/HUC6-020302.tif")
# summary(test)
```

# Watershed creation with `r.watershed`

Source: Brendan Harmon's tutorial (https://baharmon.github.io/watersheds-in-grass) with modifications for GRASS 8.3

Create hillshade (optional):
```{r}
# stringexecGRASS("r.relief input=DEM_all@PERMANENT output=Relief zscale=3")
# stringexecGRASS("r.shade shade=Relief@PERMANENT color=DEM_all@PERMANENT output=shaded_relief")
```

Make a test watershed map at a coarse scale:
```{r}
# for (i in 1:nrow(hucs.mask)){
#   # Import mask raster
#   stringexecGRASS(paste0(paste0(paste0(paste0(paste0(paste0("r.import --overwrite resample=nearest input='", projectpath), "HUC4-"), hucs.mask[i]$huc4), ".tif' output="), "HUC4_"), hucs.mask[i]$huc4))
#   
#   # Mask 
#   stringexecGRASS("r.mask -r") # remove any existing mask
#   stringexecGRASS(paste0(paste0("r.mask raster=HUC4_", hucs.mask[i]$huc4), "@PERMANENT")) # set new mask
# 
#   # Set computational region
#   stringexecGRASS(paste0(paste0("g.region raster=HUC4_", hucs.mask[i]$huc4), "@PERMANENT"))
#   
#   # Calculate watersheds
#   stringexecGRASS(paste0(paste0("r.watershed --overwrite elevation=Elevation_HUC6_30m@PERMANENT threshold=400 basin=hucs_", hucs.mask[i]$huc4), "_400 memory=2100") )
#   
#   stringexecGRASS(paste0(paste0(paste0(paste0("r.to.vect -- overwrite input=hucs_", hucs.mask[i]$huc4), "_400@PERMANENT output=hucs_"), hucs.mask[i]$huc4), "_400polys type=area")) # 200 is very small: smaller than HUC14
#   
#   # Export shapefiles
#   stringexecGRASS(paste0(paste0(paste0(paste0(paste0(paste0("v.out.ogr -e -m --overwrite input=hucs_", hucs.mask[i]$huc4), "_400polys@PERMANENT type=area output='"), projectpath), "hucs_"), hucs.mask[i]$huc4), "_400polys.shp' format=ESRI_Shapefile")) # export as multifeatures (-m), use ESRI-style .prj format (-e)
#   
#   }
```


HUCs
```{r}
# Import mask raster
stringexecGRASS(paste0(paste0("r.import --overwrite resample=nearest input='", projectpath), "HUC10-100km.tif' output=HUC10_100km"))
  
# Mask 
stringexecGRASS("r.mask -r") # remove any existing mask
stringexecGRASS("r.mask raster=HUC10_100km@PERMANENT") # set new mask

# Set computational region
stringexecGRASS("g.region raster=HUC10_100km@PERMANENT")
```

Smallest Maine HUC12 (ignoring islands) is 6.42km2, which would equal approximately 7133 raster cells. Smallest HUC14 is 1.51km2, or 1677 raster cells. So the minimum threshold size for the smaller 
```{r}
hucs_12.1 <- st_read("WBD_01_HU2_GDB.gdb", layer="WBDHU12") %>% st_transform(.,crs=st_crs(extent))
huc12 <- hucs_12.1[which(!grepl("-Atlantic Ocean",hucs_12.1$name)),] # ignore islands
# hucs_nh12.1 <- st_read("NHD_H_New_Hampshire_State_GDB.gdb", layer="WBDHU12") %>% st_transform(.,crs=st_crs(extent))
# hucs_nh12 <- hucs_nh12.1[which(grepl("-Atlantic Ocean",hucs_me12.1$name) | grepl("Island", hucs_me12.1$name)),] # ignore islands
# huc12_MENH <- dplyr::bind_rows(list(hucs_me12,hucs_nh12)) %>% st_union(.,by_feature=T) %>% st_as_sf(.)
plot(huc12["name"])
hucs_me14.1 <- st_read("NHD_H_Maine_State_GDB.gdb", layer="WBDHU14") %>% st_transform(.,crs=st_crs(extent)) # includes NH
huc12minarea <- min(huc12$areasqkm)
huc14minarea <- min(hucs_me14.1$areasqkm)

huc12meanarea <- mean(huc12$areasqkm) # corresponds closest to threshold 12800 in Friedrichs-Manthey et al. 2020
huc14meanarea <- mean(hucs_me14.1$areasqkm) # between thresholds 1600 and 3200 in above paper (using 90mx90m cells)

cellarea <- 30*30 # 1 elevation raster cell in m2
m2perkm2 <- 1000*1000 # conversion from km2 to m2

(huc14minarea*m2perkm2)/cellarea # min threshold value about 1677 cells
(huc12minarea*m2perkm2)/cellarea # min threshold value about 7133 cells

```



```{r}
scale <- 14400 # minimum number of cells required to initiate a river

# Calculate watersheds
stringexecGRASS(paste0(paste0(paste0(paste0("r.watershed -s -b --overwrite elevation=Elevation_HUC6_30m@PERMANENT threshold=", scale), " basin=hucs_100km_"), scale), "sb slope_steepness= hucs_100km_steepness memory=2100")) # D8 flow like in ArcGIS Pro (-s), beautify flat areas (-b)

# Vectorize watersheds
stringexecGRASS(paste0(paste0(paste0(paste0("r.to.vect -s --overwrite input=hucs_100km_", scale), "b@PERMANENT output=hucs_100km_"), scale), "polysbs type=area")) # 200 is very small: smaller than HUC14, smooth corners (-s)

# Export shapefiles
stringexecGRASS(paste0(paste0(paste0(paste0(paste0(paste0("v.out.ogr -e -m --overwrite input=hucs_100km_", scale), "polysbs@PERMANENT type=area output='"), projectpath), "hucs_100km_"), scale),"polysbs.shp' format=ESRI_Shapefile")) # export as multifeatures (-m), use ESRI-style .prj format (-e)
```

Calculate the area of the new watersheds:
```{r}
wsheds <- st_read(paste0(projectpath, "hucs_100km_14400polysbs.shp"))
area <- st_area(wsheds, by_feature=T)
areasqkm <- area/m2perkm2
wsheds$areasqkm <- areasqkm
mean(areasqkm) # somewhere between fine scale and intermediate scale
sd(areasqkm)
```

Clip coastal watersheds
```{r}
range.buff50.sf <- st_as_sf(range.buff50)
ocean <- vect("./Intermediate/Marine_habitats.shp")
ocean.buff <- buffer(ocean,10000)

ocean.sf <- st_read(("./Intermediate/Marine_habitats.shp"))

huc12_clip <- huc12[range.buff50.sf,]
# huc12_clip <- huc12_clip1[ocean.sf,] %>% dplyr::select(huc12, name)
st_write(huc12_clip, "./Intermediate/huc12_coast.shp", append=F)

wsheds_clip1 <- wsheds[huc12_clip,]
wsheds_clip <- wsheds_clip1 %>% vect(.) %>% erase(.,ocean) %>% st_as_sf(.)
wsheds_clipv <- vect(wsheds_clip)
st_write(wsheds_clip, "./Intermediate/watersheds_clip.shp", append = F)

land <- erase(ocean.buff,ocean)
coast <- erase(land, wsheds_clipv) %>% st_as_sf(.) %>% st_cast(.,"POLYGON")
coast.50 <- coast[range.buff50.sf,]
 
st_write(coast.50, "./Intermediate/coastal_hucs.shp", append=F)

small_hucs <- dplyr::bind_rows(list(wsheds_clip,coast.50))

st_write(small_hucs,"./Intermediate/small_hucs.shp", append=F)
small_hucs$label <- c(paste0("huc000", 1:9), 
                      paste0("huc00", 10:99),
                      paste0("huc0", 100:999),
                      paste0("huc", 1000:nrow(small_hucs)))

large_hucs <- huc12_clip %>% dplyr::select(huc12,name,areasqkm)
large_hucs$label <- c(paste0("huc12-00", 1:9), 
                      paste0("huc12-0", 10:99),
                      paste0("huc12-", 100:nrow(large_hucs)))

st_write(small_hucs, "./Inputs_vectors/small_hucs.shp", append=F)
st_write(large_hucs, "./Inputs_vectors/large_hucs.shp", append=F)

plot(small_hucs["value"])
plot(huc12_clip["states"])
```


Missing areas?
```{r}
# wsheds_clipv <- vect(wsheds_clip)
# huc12_clipv <- vect(huc12_clip)
# 
# missing <- erase(huc12_clipv, wsheds_clipv)# parts of HUC12 that are not ocean but were not estimated by GRASS
# missing_coast <- missing[ocean,]
# 
# plot(missing)
# writeVector(missing,"./Intermediate/MissingAreas.shp", overwrite=T)
```



