---
title: "Aggregating historical and current Maine and New Hampshire Bridle Shiner records at the HUC12 and HUC14 scale"
author: "Lara Katz"
date: "2025-11-06"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(sf)
library(lubridate)
library(oce) # long/lat to UTM
```

# Get centroids

```{r eval=FALSE}
# polys <- vect("./RiverCentroidPolysNH.shp")
# polys.p <- project(polys, "epsg:4326")
# 
# centers <- centroids(polys.p, inside = TRUE)
# plot(centers)
# writeVector(centers, "RiverCentroidPointsNH.shp")
# centerstab <- as.data.frame(geom(centers)[,3:4])
# centerstab$site <- centers$GNIS_Name
# write.csv(centerstab, "RiverCentroidsNH.csv")
```

# Historical point data

## Maine 

```{r eval=FALSE}
hist.me1 <- read.csv("./BDSHistoricalRecords.csv", header = T)
hist.me1[hist.me1$Year_Est == "Unknown",]$Year_Est <- -9999
hist.me1[hist.me1$YearSamp == "Unknown",]$YearSamp <- -9999
hist.me1$Year_Est <- as.integer(hist.me1$Year_Est)
hist.me1$YearSamp <- as.integer(hist.me1$YearSamp)
hist.me1$Year <- 1
for (i in 1:nrow(hist.me1)){
hist.me1[i,"Year"] <- min(c(hist.me1[i,"Year_Est"], hist.me1[i,"YearSamp"]))
}
hist.me1$Year <- as.integer(hist.me1$Year)

hist.me1 <- hist.me1[-48,]
# hist.me1[which(hist.me1$Year >= 2011),]$Site_Name
```


```{r eval=FALSE}
hist.me <- hist.me1[hist.me1$Year < 2011,]
hist.me$Date <- lubridate::mdy(hist.me$Date)
# mdy(hist.me$Date)
hist.me <- hist.me %>% 
  dplyr::select(-Year_Est,-YearSamp,-TopUTMX,-TopUTMY,-BtmUTMX,-BtmUTMY,-LengthSection,-AveWidth) %>% 
  mutate(., x=East_new, y=North_new) %>% 
  select(-East_new,-North_new)
# hist.me[hist.me$cOccu == -9999,]$cOccu <- NA
hist.me <- hist.me[,c(1,16,2,3,4,17,18,5:8,10:15)] %>% dplyr::select(-cOccu,-CStatus)
```

## New Hampshire

```{r}
all.nh <- read.csv("./NH_BS_data_dryad.csv", header = T)
all.nh$Date <- lubridate::ymd(all.nh$Date)
all.nh[all.nh$Site_Name == 0,]$Site_Name <- NA
all.nh$Year <- as.integer(all.nh$Year)
```

```{r eval=FALSE}
hist.nh1 <- all.nh[all.nh$Year < 2009,]
unique(hist.nh1$Project)
unique(hist.nh1$Water_Body)
unique(hist.nh1$Site_Name)
hist.nh1[!is.na(hist.nh1$Site_Name) & hist.nh1$Site_Name == "Blacky's Cove",]$Site_Name <- "Blackey's Cove"
hist.nh1[!is.na(hist.nh1$Site_Name) & hist.nh1$Site_Name == "",]$Site_Name <- NA
```

```{r eval=FALSE}
colnames(hist.nh1)
hist.nh <- hist.nh1 %>% 
  select(ACT_ID,Date,Site_Name,Year,Water_Body,Lat_Start,Long_Start,Survey_location_description,Survey_comments,BS) %>% 
  mutate(.,Date = Date, 
         Year = Year, 
         Water_Body=Water_Body,
         Site_Name=Site_Name,
         Site_code=ACT_ID,
         SiteHist=Survey_comments,
         x=Long_Start,
         y=Lat_Start,
         BS_Pres=BS,
         PointType = Survey_location_description) %>% 
  select(-ACT_ID,-Lat_Start,-Long_Start)
head(hist.nh)
```

```{r eval=FALSE}
hist.nh[hist.nh$BS_Pres == "P",]$BS_Pres <- "Historical"
hist.nh$HStatus <- NA
hist.nh$hOccu <- NA

hist.nh[hist.nh$BS_Pres == "Historical",]$HStatus <- "Present"

hist.nh[hist.nh$BS_Pres == "Historical",]$hOccu <- 1
hist.nh <- hist.nh[!is.na(hist.nh$x),]
utm <- lonlat2utm(hist.nh$x, hist.nh$y, zone = 19, km= FALSE)

hist.nh$x <- utm$easting
hist.nh$y <- utm$northing
hist.nh$Source <- "New Hampshire Fish & Game"
hist.nh$WBType <- NA
hist.nh$State <- "NH"
hist.nh[which(grepl("Pond", hist.nh$Water_Body)),]$WBType <- "LakePond"
hist.nh[which(grepl("Lake", hist.nh$Water_Body)),]$WBType <- "LakePond"
hist.nh[which(grepl("River", hist.nh$Water_Body)),]$WBType <- "StreamRiver"
hist.nh[which(grepl("Brook", hist.nh$Water_Body)),]$WBType <- "StreamRiver"
hist.nh[which(grepl("stream", hist.nh$Water_Body)),]$WBType <- "StreamRiver"
any(is.na(hist.nh$WBType))
colnames(hist.nh)
hist.nh <- hist.nh[,c(16,3,1,17,4,10,11,18,2,8,9,12,14,15,13)] # same order as Maine data
```

## Combined

```{r eval=FALSE}
hist.all <- rbind(hist.me,hist.nh)
hist.all[hist.all$x == -9999,] <- NA # remove points with no coordinates
hist.all <- hist.all[!is.na(hist.all$x),]
# hist.all <- transform(hist.all,
#                       WBType = as.factor(WBType),
#                       State = as.factor(State),
#                       BS_Pres = as.factor(BS_Pres),
#                       HStatus = as.factor(HStatus))

hist.all[which(grepl("19380616-Harrington",hist.all$Site_code)),]$HStatus <- "Present in vicinity of coordinates"
hist.all[which(grepl("19460618-Harrington",hist.all$Site_code)),]$HStatus <- "Present in vicinity of coordinates"
hist.all[which(grepl("1947-Canobie Lake",hist.all$Site_code)),]$HStatus <- "Present in waterbody"
hist.all[which(grepl("1947-Lower Suncook Lake",hist.all$Site_code)),]$HStatus <- "Present in waterbody"
hist.all[which(grepl("1947-Shadow Lake",hist.all$Site_code)),]$HStatus <- "Present in waterbody"
hist.all[which(grepl("1947-Wheelwright Pond",hist.all$Site_code)),]$HStatus <- "Present in waterbody"
hist.all[which(grepl("1947-Winnisquam Lake",hist.all$Site_code)),]$HStatus <- "Present in waterbody"
hist.all[which(grepl("1947-Pleasant Lake",hist.all$Site_code)),]$HStatus <- "Present in vicinity of coordinates"
hist.all[which(grepl("Fish Cove",hist.all$Site_code)),]$HStatus <- "Present in vicinity of coordinates"
hist.all[which(grepl("Mad River",hist.all$Site_code)),]$HStatus <- "Present in vicinity of coordinates"
hist.all[which(grepl("Piscataquog River",hist.all$Site_code)),]$HStatus <- "Present in vicinity of coordinates"
hist.all[which(grepl("1985",hist.all$Site_code)),]$HStatus <- "Present in vicinity of coordinates"
hist.all[which(grepl("1986",hist.all$Site_code)),]$HStatus <- "Present in vicinity of coordinates"
hist.all[which(grepl("1987",hist.all$Site_code)),]$HStatus <- "Present in vicinity of coordinates"
hist.all[which(grepl("1998",hist.all$Site_code)),]$HStatus <- "Present in vicinity of coordinates"
hist.all[which(grepl("2000",hist.all$Site_code)),]$HStatus <- "Present in vicinity of coordinates"
hist.all[which(grepl("2001",hist.all$Site_code)),]$HStatus <- "Present in vicinity of coordinates"
hist.all[which(grepl("2003",hist.all$Site_code)),]$HStatus <- "Present in vicinity of coordinates"
hist.all[which(grepl("19460618-Harrington",hist.all$Site_code)),]$Source <- "Harrington 1947: The Early Life History of the Bridled Shiner"
hist.all[which(grepl("19380616-Harrington",hist.all$Site_code)),]$Source <- "Harrington 1948: The Life Cycle and Fertility of the Bridled Shiner"
hist.all[!is.na(hist.all$PointType) & hist.all$PointType == 0,]$PointType <- NA
hist.all[!is.na(hist.all$SiteHist) & hist.all$SiteHist == 0,]$SiteHist <- "No survey comments"
hist.all[!is.na(hist.all$SiteHist) & hist.all$SiteHist == "",]$SiteHist <- "No survey comments"
# hist.all[is.na(hist.all$SiteHist),]$SiteHist <- "No survey comments"
# hist.all[is.na(hist.all$Year),]$Year <- -9999


hist.all <- hist.all[hist.all$Water_Body != "Warner River",] # remove Warner River location estimate
hist.all <- hist.all[is.na(hist.all$Site_Name) | hist.all$Site_Name != "Downstream of Suncook Lakes",]



hist.all[is.na(hist.all$PointType),]$PointType <- "GPS survey coordinates"
hist.all[hist.all$PointType == "Survey coordinates",]$PointType <- "GPS survey coordinates"
hist.all[!is.na(hist.all$Site_Name) & hist.all$Site_Name == "Mill Pond",]$PointType <- "Point in approximate center of impoundment: exact location unknown"
hist.all[!is.na(hist.all$Site_Name) & hist.all$Site_Name == "Channel between the Suncook Ponds",]$PointType <- "Point in approximate center of channel: exact location unknown"
hist.all$PointType <- as.factor(hist.all$PointType)
hist.all <- hist.all[order(hist.all$Date),]
# levels(hist.all$PointType)
summary(hist.all)
# colnames(hist.all)
hist.all <- hist.all %>% 
  rename(WaterBody = Water_Body, 
         SiteName = Site_Name, 
         SiteCode = Site_code)
```

USGS place name check (reference: GNIS database):
```{r}
hist.all[which(hist.all$SiteCode == "20080717-830-EBTJV-unnamed stream-Madison-EFISH"),]$WaterBody <- "Banfield Brook"
# hist.all[which(hist.all$WaterBody == "Brown's Brook"),]$WaterBody <- "Browns Brook"
hist.all[which(hist.all$SiteName == "GWORKB"),]$WaterBody <- "Bauneg Beg Pond"
hist.all[which(hist.all$WaterBody == "Jordan River"),]$WaterBody <- "Panther Run"
hist.all[which(hist.all$WaterBody == "Josie's Brook"),]$WaterBody <- "Josies Brook"
hist.all[which(hist.all$WaterBody == "Pemigewasset Lake"),]$WaterBody <- "Pemingewasset Lake"
hist.all[which(hist.all$WaterBody == "Saco River (Old Course)"),]$WaterBody <- "Old Course Saco River"
hist.all[which(hist.all$WaterBody == "Shaker Brook"),]$WaterBody <- "Shaker Branch"
hist.all[which(hist.all$WaterBody == "Tenny's River"),]$WaterBody <- "Tenny River"
hist.all[which(hist.all$WaterBody == "Unnamed Brook"),]$WaterBody <- "Unnamed brook"
hist.all[which(hist.all$SiteCode == "20050803-1345-WAP-Cocheco River-Farmington-SEINE"),]$WaterBody <- "Waldron Mill Pond"
hist.all[which(hist.all$SiteName == "WATBRK"),]$WaterBody <- "Unnamed brook"
hist.all[which(hist.all$WaterBody == "Winnipesaukee Lake"),]$WaterBody <- "Lake Winnipesaukee"
hist.all[which(hist.all$WaterBody == "Winnisquam Lake"),]$WaterBody <- "Lake Winnisquam"
```

Clean up years:
```{r}
hist.all[which(hist.all$WaterBody == "Canobie Lake"),]$Year <- -9999
hist.all[which(hist.all$WaterBody == "Heads Pond"),]$Year <- -9999
hist.all[which(hist.all$WaterBody == "Lamprey River" & grepl("Bunker Pond", hist.all$SiteName)),]$Year <- -9999
hist.all[which(hist.all$WaterBody == "Lower Suncook Lake"),]$Year <- -9999
hist.all[which(hist.all$WaterBody == "Pleasant Lake"),]$Year <- -9999
hist.all[which(hist.all$WaterBody == "Shadow Lake"),]$Year <- -9999
hist.all[which(hist.all$WaterBody == "Lake Winnisquam"),]$Year <- -9999
```

Clean up sources:
```{r}
hist.all[which(hist.all$Year == 1898 & hist.all$SiteName == "SEBAGO"),]$Source <- "Cooper 1939, M. Gallagher, MDIFW, written communication, 25 January 2021"
hist.all[which(hist.all$Year == 1938 & hist.all$SiteName == "BARKER"),]$Source <- "Cooper 1939, M. Gallagher, MDIFW, written communication, 25 January 2021"
hist.all[which(hist.all$Year == 1939 & hist.all$SiteName == "BEARPD"),]$Source <- "Cooper 1939, M. Gallagher, MDIFW, written communication, 25 January 2021"
hist.all[which(hist.all$Year == 1939 & hist.all$SiteName == "BURNPD"),]$Source <- "Cooper 1939, M. Gallagher, MDIFW, written communication, 25 January 2021"
hist.all[which(hist.all$Year == 1938 & hist.all$SiteName == "HIGHLK"),]$Source <- "Cooper 1939, M. Gallagher, MDIFW, written communication, 25 January 2021"
hist.all[which(hist.all$Year == 1938 & hist.all$SiteName == "STANPD"),]$Source <- "Cooper 1939, M. Gallagher, MDIFW, written communication, 25 January 2021"
hist.all[which(hist.all$Year == 1938 & hist.all$SiteName == "TRAFPD"),]$Source <- "Cooper 1939, M. Gallagher, MDIFW, written communication, 25 January 2021"
hist.all[which(hist.all$Year == 1946 & hist.all$WaterBody == "Oyster River"),]$Source <- "Harrington 1947b, NHFGD 2015 SWG Report"
hist.all[which(hist.all$Year == 1947 & hist.all$WaterBody == "Wheelwright Pond"),]$Source <- "Harrington 1947b, NHFGD 2015 SWG Report"
hist.all[which(hist.all$Year == 1946 & hist.all$WaterBody == "Oyster River"),]$Source <- "Harrington 1947b, NHFGD 2015 SWG Report"
hist.all[which(hist.all$Year == 1938 & hist.all$WaterBody == "Suncook River"),]$Source <- "Harrington 1948"
hist.all[which(hist.all$Year == 1906 & hist.all$WaterBody == "Tenny River"),]$Source <- "Kendall 1914"
hist.all[which(hist.all$Year == 1899 & hist.all$WaterBody == "Chaffin Pond"),]$Source <- "Kendall 1914"
hist.all[which(hist.all$SiteName == "CRESLK-02"),]$Source <- "Kendall 1914"
hist.all[which(hist.all$Year == 1898 & hist.all$WaterBody == "Little Sebago Lake"),]$Source <- "Kendall 1914"
hist.all[which(hist.all$WaterBody == "Sebago Lake" & hist.all$SiteName == "Kettle Cove"),]$Source <- "Kendall 1914"
hist.all[which(hist.all$SiteName == "SEBAGO-01"),]$Source <- "Kendall 1914, M. Gallagher, MDIFW, written communication, 25 January 2021"
hist.all[which(hist.all$Source == "New Hampshire Fish & Game"),]$Source <- "M. Carpenter, NHFGD, written communication, 7 February 2023"
hist.all[which(hist.all$Source == "Maine Department of Inland Fisheries & Wildlife Records"),]$Source <- "M. Gallagher, MDIFW, written communication, 25 January 2021"
hist.all[which(hist.all$SiteName == "PROCPD"),]$Source <- "USEPA 2016"
hist.all[which(hist.all$Year == -9999 & hist.all$WaterBody == "Heads Pond"),]$Source <- "NHFGD 2015 SWG Report"
hist.all[which(hist.all$Year == -9999 & hist.all$WaterBody == "Lamprey River"),]$Source <- "NHFGD 2015 SWG Report"
hist.all[which(hist.all$Year == -9999 & hist.all$WaterBody == "Canobie Lake"),]$Source <- "NHFGD 2015 SWG Report"
hist.all[which(hist.all$Year == -9999 & hist.all$WaterBody == "Lower Suncook Lake"),]$Source <- "NHFGD 2015 SWG Report"
hist.all[which(hist.all$Year == -9999 & hist.all$WaterBody == "Pleasant Lake"),]$Source <- "NHFGD 2015 SWG Report"
hist.all[which(hist.all$Year == -9999 & hist.all$WaterBody == "Shadow Lake"),]$Source <- "NHFGD 2015 SWG Report"
hist.all[which(hist.all$Year == -9999 & hist.all$WaterBody == "Lake Winnisquam"),]$Source <- "NHFGD 2015 SWG Report"
hist.all[which(hist.all$PointType == "Digitized from NH Fish survey map (StoryMap)"),]$Source <- "NH Fish StoryMap"
hist.all[which(hist.all$SiteName == "BROBRK"),]$Source <- "M. Gallagher, MDIFW, written communication, 25 January 2021"
```

Clean up site names:
```{r}
hist.all[which(hist.all$State == "NH"),]$SiteName <- hist.all[which(hist.all$State == "NH"),]$SiteCode
hist.all[which(hist.all$State == "ME" & hist.all$SiteName == ""),]$SiteName <- hist.all[which(hist.all$State == "ME" & hist.all$SiteName == ""),]$WaterBody
```

Clean up point type:
```{r}
hist.all[which(hist.all$SiteName == "20060626-1030-WAP-Lamprey River-Raymond-SEINE"),]$PointType <- "GPS survey coordinates"
hist.all[which(hist.all$WaterBody == "Banfield Brook"),]$PointType <- "GPS survey coordinates"
hist.all[which(hist.all$SiteName == "20050812-1200-WAP-Suncook River-Pembroke-SEINE"),]$PointType <- "GPS survey coordinates"
hist.all[which(hist.all$SiteName == "20060626-1155-WAP-Exeter River-Fremont-SEINE"),]$PointType <- "GPS survey coordinates"
hist.all[which(hist.all$SiteName == "20080521-1100-HERB-Pemigewasset Lake-New Hampton-EBOAT"),]$PointType <- "GPS survey coordinates"
hist.all[which(hist.all$PointType == "Lake centroid"),]$PointType <- "Center of waterbody: no precise location known"
hist.all[which(hist.all$PointType == "Pond centroid"),]$PointType <- "Center of waterbody: no precise location known"
hist.all[which(hist.all$PointType == "Unknown survey location: lat/long point in center of pond"),]$PointType <- "Center of waterbody: no precise location known"
```

### Save outputs

```{r warning=FALSE}
# write.csv(hist.all, "./HistoricalBridleShinerPoints.csv")
hist <- st_as_sf(hist.all, coords=c("x","y"), crs=26919, 
                    remove = F, # remove x and y columns from data.frame
                    agr="identity" # columns (attributes) uniquely identify the geometry of each point
                    )
# write_sf(hist,"./HistoricalBridleShinerPoints.shp",overwrite=FALSE)
```


# Current point data

## Maine

```{r}
curr.me1 <- read.csv("./BDS_all_points_current.csv", header = T)

curr.me1 <- curr.me1 %>% filter(BDS_PA != -9999) %>% filter(Sample != "Control")
curr.me1[curr.me1$WaterBody == "\"Robert's Ridge\" brook",]$WaterBody <- "Unnamed brook" 
curr.me1[curr.me1$WaterBody == "Canco Woods brook",]$WaterBody <- "Unnamed brook"
curr.me1[curr.me1$WaterBody == "Watchic Brook",]$WaterBody <- "Watchic Pond outlet brook"
curr.me1[curr.me1$WaterBody == "\"Watchic Brook\"",]$WaterBody <- "Watchic Pond outlet brook"
curr.me1[curr.me1$WaterBody == "Great Works River/Bauneg Beg Pond",]$WaterBody <- "Bauneg Beg Pond"
curr.me1[which(grepl("Boom Rd",curr.me1$WaterBody)),]$WaterBody <- "Unnamed brook"
curr.me1[curr.me1$WaterBody == "Brown's Brook",]$WaterBody <- "Browns Brook"
curr.me1[curr.me1$SiteName == "PRESUM",]$SiteName <- "PRESUM-01" 
curr.me1[curr.me1$SiteName == "TRAFPD" & curr.me1$Subsite == "A",]$SiteName <- "TRAFPD-01" 
curr.me1[curr.me1$SiteName == "TRAFPD" & curr.me1$Subsite == "B",]$SiteName <- "TRAFPD-02"

curr.me1$WaterBody <- as.factor(curr.me1$WaterBody)
curr.me1$SiteName <- as.factor(curr.me1$SiteName)
levels(curr.me1$SiteName)
curr.me1$BDS_PA <- as.integer(curr.me1$BDS_PA)
curr.me1$PointGroup <- curr.me1$SiteName # placeholder
```

```{r warning=FALSE}
curr.me2 <- st_as_sf(curr.me1, coords=c("Easting","Northing"), crs=26919, 
                    remove = F, # remove x and y columns from data.frame
                    agr="identity" # columns (attributes) uniquely identify the geometry of each point
                    )

# write_sf(curr.me2,"./CurrentPointsUngrouped.shp", overwrite = FALSE)
```

```{r}
curr.gr1 <- st_read("./CurrentPointsGrouped.shp")
curr.gr1$PontGrp <- as.factor(curr.gr1$PontGrp)
# curr.gr1$WatrBdy <- as.factor(curr.gr1$WatrBdy)
curr.me.join <- curr.me1 %>% select(SiteName,Date) %>% arrange(as_date(mdy(Date)))

curr.gr2 <- st_drop_geometry(curr.gr1) %>%
  select(PontGrp, SiteNam, WatrBdy, Easting, Northng, BDS_PA, SmplngN) %>% mutate(
    PointGroup = PontGrp,
    SiteName = SiteNam,
    WaterBody = WatrBdy,
    Northing = Northng,
    SiteHist = SmplngN) %>% 
  select(-PontGrp, -SiteNam, -WatrBdy, -Northng, -SmplngN) %>% 
  left_join(., curr.me.join, by = "SiteName", multiple = "first")

curr.gr2[curr.gr2$SiteName == "GWORKB",]$WaterBody = "Bauneg Beg Pond"
curr.gr2[curr.gr2$SiteName == "SACONO-02",]$WaterBody = "Saco River impoundment"
curr.gr2[curr.gr2$SiteName == "SACONO-03",]$WaterBody = "Saco River impoundment"
curr.gr2[curr.gr2$SiteName == "HIGHLK-03",]$WaterBody = "Highland Lake outlet (Stevens Brook)"
curr.gr2[curr.gr2$PointGroup == "RANGE-01-02",]$WaterBody = "Middle Range Pond inlet (unnamed pond)"
curr.meanpts <- curr.gr2 %>% group_by(PointGroup) %>% arrange(as_date(mdy(Date))) %>% 
  summarise(SiteName=first(SiteName),
            WaterBody = first(WaterBody),
            Mean.x = mean(Easting),
            Mean.y = mean(Northing),
            PA = max(BDS_PA),
            Date=min(as_date(mdy(Date))), # earliest date surveyed
            .groups = 'drop')

curr.meanpts.sf <- st_as_sf(curr.meanpts, coords=c("Mean.x","Mean.y"), crs=26919, 
                    remove = F, # remove x and y columns from data.frame
                    agr="identity" # columns (attributes) uniquely identify the geometry of each point
                    )
# write_sf(curr.meanpts.sf,"./CurrentMeanPointsME.shp", overwrite = FALSE)
```

## New Hampshire

```{r}
curr.nh2 <- all.nh[all.nh$Year >= 2009,]

unique(curr.nh2[order(curr.nh2$Water_Body),]$Water_Body)
curr.nh2[!is.na(curr.nh2$Water_Body) & curr.nh2$Water_Body == "Lake Winnipesaukee",]$Water_Body <- "Winnipesaukee Lake"
curr.nh2[curr.nh2$Water_Body == "Harper Brook  ",]$Water_Body <- "Harper Brook"

unique(curr.nh2[order(curr.nh2$Site_Name),]$Site_Name)
curr.nh2[!is.na(curr.nh2$Site_Name) & curr.nh2$Site_Name == "Blackeys Cove",]$Site_Name <- "Blackey's Cove"
curr.nh2[!is.na(curr.nh2$Site_Name) & curr.nh2$Site_Name == "Greens Basin",]$Site_Name <- "Green's Basin"
curr.nh2[!is.na(curr.nh2$Site_Name) & curr.nh2$Site_Name == "Lees Mills YOY Bass",]$Site_Name <- "Lees Mills"
curr.nh2[!is.na(curr.nh2$Site_Name) & curr.nh2$Site_Name == "",]$Site_Name <- NA
```

```{r}
# colnames(curr.nh2)
curr.nh1 <- curr.nh2 %>% 
  select(ACT_ID,Date,Site_Name,Year,Water_Body,Lat_Start,Long_Start,Survey_location_description,Survey_comments,BS) %>% 
  mutate(.,
         Date = Date, 
         Year = Year, 
         Water_Body=Water_Body,
         Site_Name=Site_Name,
         Site_code=ACT_ID,
         SiteHist=Survey_comments,
         x=Long_Start,
         y=Lat_Start,
         BS_Pres=BS,
         PointType = "GPS survey coordinates") %>% 
  select(-ACT_ID,-Lat_Start,-Long_Start)
head(curr.nh1)
```

```{r}
curr.nh1[curr.nh1$BS_Pres == "P",]$BS_Pres <- "Current"
curr.nh1[curr.nh1$BS_Pres == "A",]$BS_Pres <- "Absent"
curr.nh1$CStatus <- NA
curr.nh1$cOccu <- NA

curr.nh1[curr.nh1$BS_Pres == "Current",]$CStatus <- "Present"
curr.nh1[curr.nh1$BS_Pres == "Absent",]$CStatus <- "Absent"
curr.nh1[curr.nh1$BS_Pres == "Current",]$cOccu <- 1
curr.nh1[curr.nh1$BS_Pres == "Absent",]$cOccu <- 0


curr.nh1 <- curr.nh1[!is.na(curr.nh1$x),]
utm <- lonlat2utm(curr.nh1$x, curr.nh1$y, zone = 19, km= FALSE)

curr.nh1$x <- utm$easting
curr.nh1$y <- utm$northing
curr.nh1$Source <- "New Hampshire Fish & Game"
curr.nh1$WBType <- NA
curr.nh1$State <- "NH"
curr.nh1[which(grepl("Pond", curr.nh1$Water_Body)),]$WBType <- "LakePond"
curr.nh1[which(grepl("pond", curr.nh1$Water_Body)),]$WBType <- "LakePond"
curr.nh1[which(grepl("Lake", curr.nh1$Water_Body)),]$WBType <- "LakePond"
curr.nh1[which(grepl("River", curr.nh1$Water_Body)),]$WBType <- "StreamRiver"
curr.nh1[which(grepl("Brook", curr.nh1$Water_Body)),]$WBType <- "StreamRiver"
curr.nh1[which(grepl("stream", curr.nh1$Water_Body)),]$WBType <- "StreamRiver"
any(is.na(curr.nh1$WBType))
curr.nh1 <- curr.nh1[,c(16,3,1,17,4,10,11,18,2,8,9,12,14,15,13)]
```

```{r}
curr.nh1[!is.na(curr.nh1$Site_Name) & curr.nh1$Site_Name == "Mill Pond",]$CStatus <- "Presumed extirpated"
curr.nh1[curr.nh1$Water_Body == "Canobie Lake",]$CStatus <- "Extirpated"
curr.nh1[which(grepl("UNK-Lower Suncook Lake",curr.nh1$Site_code)),]$CStatus <- "Presumed extirpated"
curr.nh1[which(grepl("UNK-Pleasant Lake",curr.nh1$Site_code)),]$CStatus <- "Extirpated"
curr.nh1[curr.nh1$Water_Body == "Shadow Lake",]$CStatus <- "Extirpated"
curr.nh1[curr.nh1$Water_Body == "Wheelwright Pond",]$CStatus <- "Presumed extirpated"
curr.nh1[curr.nh1$Water_Body == "Winnisquam Lake",]$CStatus <- "Extirpated"
curr.nh1[curr.nh1$Water_Body == "Heads Pond",]$CStatus <- "Extirpated"
curr.nh1[!is.na(curr.nh1$Site_Name) & curr.nh1$Site_Name == "Bunker Pond",]$CStatus <- "Extirpated"

curr.nh1[curr.nh1$SiteHist == "2015 Fish Survey Presence Absence Data",]$PointType <- "Digitized from NH Fish survey map (StoryMap)"

curr.nh1[curr.nh1$Year == 9999 & is.na(curr.nh1$Site_Name),]$PointType <- "Center of waterbody: no precise location known"

curr.nh1[!is.na(curr.nh1$Site_Name) & curr.nh1$Site_Name == "Fish Cove",]$PointType <- "Digitized coordinates: northern end of cove"

curr.nh1[!is.na(curr.nh1$Site_Name) & curr.nh1$Site_Name == "Bunker Pond",]$PointType <- "Digitized coordinates from 1992 aerial imagery (Google Earth)"

curr.nh1[!is.na(curr.nh1$Site_Name) & curr.nh1$Site_Name == "Mill Pond",]$PointType <- "Point in approximate center of impoundment: exact location unknown"

curr.nh1 <- transform(curr.nh1,
                      WBType = as.factor(WBType),
                      BS_Pres = as.factor(BS_Pres),
                      CStatus = as.factor(CStatus))

curr.nh1$PointGroup <- curr.nh1$Site_code
curr.nh <- st_as_sf(curr.nh1, coords=c("x","y"), crs=26919, 
                    remove = F, # remove x and y columns from data.frame
                    agr="identity" # columns (attributes) uniquely identify the geometry of each point
                    )

# write_sf(curr.nh,"./CurrentPointsUngroupedNH.shp", overwrite = FALSE)
```

```{r}
curr.grnh1 <- st_read("./CurrentPointsGroupedNH2.shp")
curr.grnh1[curr.grnh1$PointGroup == "Branch River-Milton",]$PointGroup <- "BS-Branch River-Milton"
curr.grnh1 <- curr.grnh1[order(curr.grnh1$PointGroup),]
curr.grnh1[which(grepl("^BS-",curr.grnh1$PointGroup)),]$PointGroup <- paste0(curr.grnh1[which(grepl("^BS-",curr.grnh1$PointGroup)),]$PointGroup,"-MEANCENTER")
# curr.grnh1[which(grepl("^BS-",curr.grnh1$PointGroup)),]$PointGroup # CHECK


curr.grnh1$PointGroup <- as.factor(curr.grnh1$PointGroup)
curr.grnh1$Water_Body <- as.factor(curr.grnh1$Water_Body)
curr.grnh2 <- st_drop_geometry(curr.grnh1) %>% 
  select(PointGroup,Site_code,Water_Body,x,y,cOccu) %>% 
  group_by(PointGroup) %>% 
  summarise(SiteName=first(Site_code), 
            WaterBody = first(Water_Body), 
            Mean.x = mean(x), 
            Mean.y = mean(y), 
            PA = max(cOccu), 
            .groups = 'drop')

curr.grnh <- st_as_sf(curr.grnh2, coords=c("Mean.x","Mean.y"), crs=26919, 
                    remove = F, # remove x and y columns from data.frame
                    agr="identity" # columns (attributes) uniquely identify the geometry of each point
                    )
# write_sf(curr.grnh,"./CurrentMeanPointsNH.shp", overwrite = FALSE)
```

## Combined

```{r}
current.me.sf <- st_read("./CurrentMeanPointsME.shp")
current.nh.sf <- st_read("./CurrentMeanPointsNH.shp")
```

```{r}
current.me.df <- st_drop_geometry(current.me.sf)
current.me.df$State <- "ME"

current.nh.df <- st_drop_geometry(current.nh.sf)
current.nh.df$Date <- NA # placeholder
current.nh.df$State <- "NH"

current.all.df1 <- rbind(current.me.df,current.nh.df)

colnames(current.all.df1)[6] <- "cOccu"
# colnames(current.all.df1)[1] <- "Site_code"

# mejoin <- curr.me1 %>% select(Date,PointGroup) %>% group_by(PointGroup) %>% summarise(Date=min(Date))
nhjoin <- curr.nh1 %>% select(-Water_Body,-x,-y,-State,-Site_Name,-cOccu) 

current.all.df2 <- left_join(current.all.df1,nhjoin,by="PointGroup")
# left_join(current.all.df1,mejoin, by = "PointGroup") %>%
```

Clean up
```{r}
# Fill in dates for NH data
current.all.df2[current.all.df2$State == "NH" & 
                  is.na(current.all.df2$Date.x),]$Date.x <- current.all.df2[current.all.df2$State == "NH" &                                                               is.na(current.all.df2$Date.x),]$Date.y
# Remove Date.y column
curr.nh.join <- curr.grnh1 %>% select(-Water_Body,-x,-y,-State,-Site_Name,-Site_code,-cOccu) %>% st_drop_geometry(.) %>% arrange(Date)
current.all.df.date <- current.all.df2 %>% mutate(Date = Date.x) %>% select(-Date.x,-Date.y) %>% left_join(.,curr.nh.join, by="PointGroup", multiple = "first")
# current.all.df2$Date <- as_date(current.all.df2$Date)

current.all.df.date[which(grepl("-MEANCENTER",current.all.df.date$PointGroup)),]$Date.x <- current.all.df.date[which(grepl("-MEANCENTER",current.all.df.date$PointGroup)),]$Date.y
```
[1] "Source"     "Year"       "Date"       "WBType"     "Water_Body" "x"          "y"         
 [8] "State"      "Site_Name"  "Site_code"  "SiteHist"   "BS_Pres"    "HStatus"    "hOccu"     
[15] "PointType" 

```{r}
current.all.df.date[current.all.df.date$State == "NH" & 
                  is.na(current.all.df.date$Source.x),]$Source.x <- current.all.df.date[current.all.df.date$State == "NH" & 
                  is.na(current.all.df.date$Source.x),]$Source.y
current.all.df.date[current.all.df.date$State == "ME" & 
                  is.na(current.all.df.date$Source.x),]$Source.x <- "University of Maine Bridle Shiner Surveys 2021-2022"
```

Year
```{r}
current.all.df.date[is.na(current.all.df.date$Year.x),]$Year.x <- year(current.all.df.date[is.na(current.all.df.date$Year.x),]$Date.x)
```

Water body type
```{r}
current.all.df.date[current.all.df.date$State == "NH" & 
                  is.na(current.all.df.date$WBType.x),]$WBType.x <- current.all.df.date[current.all.df.date$State == "NH" & 
                  is.na(current.all.df.date$WBType.x),]$WBType.y
current.all.df.date$WBType.x <- as.character(current.all.df.date$WBType.x)

current.all.df.date$WBType <- NA

current.all.df.date[which(grepl("Pond", current.all.df.date$WaterBody)),]$WBType <- "LakePond"
current.all.df.date[which(grepl("Lake", current.all.df.date$WaterBody)),]$WBType <- "LakePond"
current.all.df.date[which(grepl("Pond", current.all.df.date$WaterBody)),]$WBType <- "LakePond"
current.all.df.date[which(grepl("River", current.all.df.date$WaterBody)),]$WBType <- "StreamRiver"
current.all.df.date[which(grepl("Brook", current.all.df.date$WaterBody)),]$WBType <- "StreamRiver"
current.all.df.date[which(grepl("Unnamed brook", current.all.df.date$WaterBody)),]$WBType <- "StreamRiver"
current.all.df.date[current.all.df.date$WaterBody == "The Heath",]$WBType <- "LakePond"
current.all.df.date[which(grepl("unnamed pond", current.all.df.date$WaterBody)),]$WBType <- "LakePond"
current.all.df.date[which(grepl("outlet brook", current.all.df.date$WaterBody)),]$WBType <- "StreamRiver"
current.all.df.date[current.all.df.date$SiteName == "SACONO-03",]$WBType <- "LakePond" # no longer connected to the Saco River
current.all.df.date[current.all.df.date$State == "ME",]$WBType.x <- current.all.df.date[current.all.df.date$State == "ME",]$WBType

current.all.df.wb <- current.all.df.date %>% 
  select(-Source.y, -Year.y, -Date.y, -WBType.y, -WBType, -Site_code) %>% 
  mutate(SiteCode = PointGroup, Source = Source.x, Year = Year.x, 
         Date = Date.x, WBType = WBType.x) %>% 
  select(-PointGroup,-Source.x, -Year.x, -Date.x, -WBType.x)
```

Survey coordinates
```{r}
current.all.df.wb[which(grepl("-MEANCENTER", current.all.df.wb$SiteCode)),]$PointType.x <- "Mean center point of 2+ surveys at this location"
current.all.df.wb[current.all.df.wb$State == "ME",]$PointType.x <- "Mean center point of seine and/or eDNA sample replicates collected at this location"
current.all.df.wb[current.all.df.wb$State == "ME" & current.all.df.wb$Year == 2022,]$PointType.x <- "Mean center point of eDNA sample replicates collected at this location"

current.all.df.pt <- current.all.df.wb %>% select(-PointType.y) %>% mutate(PointType = PointType.x) %>% select(-PointType.x)
```

Site presence
```{r}
current.all.df.pt[which(grepl("-MEANCENTER", current.all.df.pt$SiteCode)),]$BS_Pres.x <- current.all.df.pt[which(grepl("-MEANCENTER", current.all.df.pt$SiteCode)),]$BS_Pres.y 
current.all.df.pt[which(grepl("-MEANCENTER", current.all.df.pt$SiteCode)),]$CStatus.x <- current.all.df.pt[which(grepl("-MEANCENTER", current.all.df.pt$SiteCode)),]$CStatus.y 

current.all.df.pt[current.all.df.pt$State == "ME" & current.all.df.pt$cOccu == 1,]$BS_Pres.x <- "Current"
current.all.df.pt[current.all.df.pt$State == "ME" & current.all.df.pt$cOccu == 1,]$CStatus.x <- "Present"

current.all.df.pt$BS_Pres.x <- as.character(current.all.df.pt$BS_Pres.x)
current.all.df.pt$CStatus.x <- as.character(current.all.df.pt$CStatus.x)

# Lake/pond sites in Maine for which we had no precise historical survey location
current.all.df.pt[current.all.df.pt$State == "ME" & 
                    current.all.df.pt$cOccu == 0 & 
                    current.all.df.pt$Year == 2021,]$BS_Pres.x <- "Historically present in waterbody"

current.all.df.pt[current.all.df.pt$State == "ME" & 
                    current.all.df.pt$cOccu == 0 & 
                    current.all.df.pt$WaterBody == "Saco River",]$BS_Pres.x <- "Historically present in the river and surrounding watershed"

current.all.df.pt[current.all.df.pt$cOccu == 0 & 
                    current.all.df.pt$WaterBody == "Ossipee River",]$BS_Pres.x <- "Historically present in the river and surrounding watershed"

# Stream/river sites for which we had historical survey coordinates
current.all.df.pt[current.all.df.pt$State == "ME" & 
                    current.all.df.pt$cOccu == 0 & 
                    current.all.df.pt$Year == 2021 &
                    current.all.df.pt$WBType == "StreamRiver" & 
                    current.all.df.pt$SiteName != "HIGHLK-03" & 
                    current.all.df.pt$WaterBody != "Tenny River",]$BS_Pres.x <- "Historically present in vicinity of coordinates"

# River sites for which we had no precise historical survey location
current.all.df.pt[current.all.df.pt$WaterBody == "Tenny River",]$BS_Pres.x <- "Historically present in waterbody"


# Lake/pond sites for which we had an approximate historical survey location
current.all.df.pt[current.all.df.pt$WaterBody == "Bauneg Beg Pond",]$BS_Pres.x <- "Historically present in vicinity of coordinates"
current.all.df.pt[current.all.df.pt$WaterBody == "Spectacle Pond",]$BS_Pres.x <- "Historically present in vicinity of coordinates"


# Sites for which we had no historical occupancy data
current.all.df.pt[current.all.df.pt$State == "ME" & 
                    current.all.df.pt$cOccu == 0 & 
                    current.all.df.pt$Year == 2022,]$BS_Pres.x <- "No historical record of occupancy"

# But then correct for Sokokis and Crescent Lake
current.all.df.pt[which(grepl("SOKOLK", current.all.df.pt$SiteName)),]$BS_Pres.x <- "Historically present in waterbody"
current.all.df.pt[which(grepl("CRESLK", current.all.df.pt$SiteName)),]$BS_Pres.x <- "Historically present in waterbody"

# Lake Winnipesaukee sites
current.all.df.pt[current.all.df.pt$WaterBody == "Winnipesaukee Lake",]$BS_Pres.x <- "Historically present in waterbody"
current.all.df.pt[which(grepl("Blackeys Cove", current.all.df.pt$SiteCode)),]$BS_Pres.x <- "Historically present in vicinity of coordinates"
current.all.df.pt[which(grepl("Fish Cove", current.all.df.pt$SiteCode)),]$BS_Pres.x <- "Historically present in vicinity of coordinates"

# NH river sites where bridle shiners were historically present somewhere within the river
current.all.df.pt[current.all.df.pt$WaterBody == "Soucook River",]$BS_Pres.x <- "Historically present in the river and surrounding watershed"
current.all.df.pt[current.all.df.pt$WaterBody == "Lamprey River",]$BS_Pres.x <- "Historically present in river"
current.all.df.pt[current.all.df.pt$WaterBody == "Suncook River",]$BS_Pres.x <- "Historically present in the river and surrounding watershed"
current.all.df.pt[current.all.df.pt$WaterBody == "Isinglass River",]$BS_Pres.x <- "Historically present in river"
current.all.df.pt[current.all.df.pt$WaterBody == "Oyster River",]$BS_Pres.x <- "Historically present in river"
current.all.df.pt[current.all.df.pt$WaterBody == "Powwow River",]$BS_Pres.x <- "Historically present in river"
current.all.df.pt[current.all.df.pt$WaterBody == "Exeter River",]$BS_Pres.x <- "Historically present in river"
current.all.df.pt[current.all.df.pt$WaterBody == "Cocheco River",]$BS_Pres.x <- "Historically present in river"

current.all.df.pt[current.all.df.pt$SiteCode == "BS-Lamprey River-Epping-MEANCENTER",]$BS_Pres.x <- "Historically present in vicinity of coordinates"




current.all.df.pt[current.all.df.pt$State == "ME" & 
                    current.all.df.pt$cOccu == 0,]$CStatus.x <- "Absent"

current.all.df.pt[current.all.df.pt$BS_Pres.x == "Absent",]$BS_Pres.x <- "No historical record of occupancy"
current.all.df.pt[current.all.df.pt$CStatus.x == "Extirpated",]$BS_Pres.x <- "Historically present in waterbody"
current.all.df.pt[current.all.df.pt$CStatus.x == "Presumed extirpated",]$BS_Pres.x <- "Historically present in waterbody"
current.all.df.pt[current.all.df.pt$SiteCode == "UNK-Lamprey River-Bunker Pond-Epping",]$BS_Pres.x <- "Historically present in vicinity of coordinates"
current.all.df.pt[current.all.df.pt$SiteCode == "UNK-Oyster River-Mill Pond-Durham",]$BS_Pres.x <- "Historically present in vicinity of coordinates"

current.all.stat <- current.all.df.pt %>% select(-BS_Pres.y,-CStatus.y,-SiteHist.x) %>% mutate(BS_Pres = BS_Pres.x, CStatus = CStatus.x, SiteHist=SiteHist.y) %>% select(-BS_Pres.x,-CStatus.x,-SiteHist.y)
```

```{r}
me.sitehist.join <- curr.gr2 %>% mutate(SiteCode = PointGroup) %>% select(SiteCode,SiteHist)
current.all <- left_join(current.all.stat, me.sitehist.join, by="SiteCode",multiple = "first")
current.all[current.all$State == "ME",]$SiteHist.x <- current.all[current.all$State == "ME",]$SiteHist.y
current.all <- current.all %>% mutate(SiteHist = SiteHist.x) %>% select(-SiteHist.x,-SiteHist.y)
current.all <- current.all[,c(8:11,2,6,3,4,1,7,15,13,14,5,12)]
current.all <- current.all[order(current.all$Date),]
```

USGS place name check (reference: GNIS database):
```{r}
current.all[which(current.all$SiteName == "20170913-1300-BS-Little Suncook River-Epsom-DIPNET"),]$WaterBody <- "Bixby Pond"

current.all[which(current.all$SiteName == "20210723-1200-BS-Berry Pond-Moultonborough-DIPNET"),]$WaterBody <- "Weed Brook"
# current.all[which(current.all$WaterBody == "Brown's Brook"),]$WaterBody <- "Browns Brook"
current.all[which(current.all$WaterBody == "Jordan River"),]$WaterBody <- "Panther Run"
current.all[which(current.all$WaterBody == "Josie's Brook"),]$WaterBody <- "Josies Brook"
current.all[which(current.all$WaterBody == "Half Moon Pond (Otter Ponds complex)"),]$WaterBody <- "Half Moon Pond"
current.all[which(current.all$WaterBody == "Heron Pond"),]$WaterBody <- "Unnamed pond"
current.all[which(current.all$WaterBody == "Harper Brook tributary"),]$WaterBody <- "Harper Brook"
current.all[which(current.all$WaterBody == "Highland Lake outlet (Stevens Brook)"),]$WaterBody <- "Highland Lake"
current.all[which(current.all$WaterBody == "Kanasatka Lake"),]$WaterBody <- "Lake Kanasatka"
current.all[which(current.all$WaterBody == "Pemigewasset Lake"),]$WaterBody <- "Pemingewasset Lake"
# current.all[which(current.all$WaterBody == "Saco River (Old Course)"),]$WaterBody <- "Old Course Saco River"
# current.all[which(current.all$WaterBody == "Shaker Brook"),]$WaterBody <- "Shaker Branch"
current.all[which(current.all$SiteName == "RACHEL"),]$WaterBody <- "Spurwink River"
current.all[which(current.all$SiteName == "RANGE-01"),]$WaterBody <- "Unnamed brook"
current.all[which(current.all$WaterBody == "Saco River impoundment"),]$WaterBody <- "Saco River"
current.all[which(current.all$SiteName == "SEBAGO-04"),]$WaterBody <- "Sticky River"
current.all[which(current.all$WaterBody == "Shepard's River"),]$WaterBody <- "Shepards River"
current.all[which(current.all$WaterBody == "Snake Pond (Otter Ponds complex)"),]$WaterBody <- "Snake Pond"
current.all[which(current.all$WaterBody == "Snake River"),]$WaterBody <- "Lake Waukewan"
# current.all[which(current.all$WaterBody == "Tenny's River"),]$WaterBody <- "Tenny River"
current.all[which(current.all$WaterBody == "Warren Hatchery Pond"),]$WaterBody <- "Unnamed pond"
current.all[which(current.all$SiteName == "WATBRK"),]$WaterBody <- "Unnamed brook"
current.all[which(current.all$WaterBody == "Wentworth Lake"),]$WaterBody <- "Lake Wentworth"
current.all[which(current.all$WaterBody == "Wickwas Lake"),]$WaterBody <- "Lake Wicwas"
current.all[which(current.all$WaterBody == "Winnipesaukee Lake"),]$WaterBody <- "Lake Winnipesaukee"
current.all[which(current.all$WaterBody == "Winnisquam Lake"),]$WaterBody <- "Lake Winnisquam"
```

Source clean up:
```{r}
current.all[which(current.all$Source == "New Hampshire Fish & Game"),]$Source <- "M. Carpenter, NHFGD, written communication, 7 February 2023"
current.all[which(current.all$WaterBody == "Connecticut River"),]$Source <- "NH Fish StoryMap"
current.all[which(current.all$Year == 9999 & current.all$WaterBody == "Canobie Lake"),]$Source <- "NHFGD 2015 SWG Report"
current.all[which(current.all$Year == 9999 & current.all$WaterBody == "Lamprey River"),]$Source <- "NHFGD 2015 SWG Report"
current.all[which(current.all$Year == 9999 & current.all$WaterBody == "Lower Suncook Lake"),]$Source <- "NHFGD 2015 SWG Report"
current.all[which(current.all$Year == 9999 & current.all$WaterBody == "Oyster River"),]$Source <- "NHFGD 2015 SWG Report"
current.all[which(current.all$Year == 9999 & current.all$WaterBody == "Pleasant Lake"),]$Source <- "NHFGD 2015 SWG Report"
current.all[which(current.all$Year == 9999 & current.all$WaterBody == "Shadow Lake"),]$Source <- "NHFGD 2015 SWG Report"
current.all[which(current.all$Year == 9999 & current.all$WaterBody == "Wheelwright Pond"),]$Source <- "NHFGD 2015 SWG Report"
current.all[which(current.all$Year == 9999 & current.all$WaterBody == "Lake Winnipesaukee"),]$Source <- "NHFGD 2015 SWG Report"
current.all[which(current.all$Year == 9999 & current.all$WaterBody == "Lake Winnisquam"),]$Source <- "NHFGD 2015 SWG Report"
```

Clean up WBType:
```{r}
current.all[which(current.all$WaterBody == "Sticky River"),]$WBType <- "StreamRiver"
```

### Save outputs

```{r}
# write.csv(current.all, "./CurrentBridleShinerPoints.csv")
current <- st_as_sf(current.all, coords=c("Mean_x","Mean_y"), crs=26919, 
                    remove = F, # remove x and y columns from data.frame
                    agr="identity" # columns (attributes) uniquely identify the geometry of each point
                    )
# write_sf(current,"./CurrentBridleShinerPoints.shp",overwrite=FALSE)
```

# Supplement D: Table SD1

```{r}
current.all <- current.all %>% 
  # arrange(-cOccu, WaterBody) %>% 
  # relocate(.,SiteName) %>% 
  # relocate(.,WaterBody,.after = SiteName) %>% 
  # relocate(.,cOccu,.after = WaterBody) %>% 
  # relocate(.,Year,.after = cOccu) %>% 
  # relocate(.,State,.after = Year) %>% 
  # relocate(.,c(Mean_x, Mean_y),.after = State) %>% 
  # relocate(.,PointType,.after = Mean_y) %>% 
  # relocate(.,Source,.after = PointType) %>% 
  # select(-Date, -WBType, -SiteCode, -SiteHist, -BS_Pres, -CStatus) %>% 
  # rename(Easting = Mean_x, Northing = Mean_y)
  rename(Occupancy = cOccu)

current.all[which(current.all$Occupancy == 1),]$Occupancy <- "Current presence"
current.all[which(current.all$Occupancy == 0),]$Occupancy <- "Current absence"
current.all[which(current.all$Year == 9999),]$Year <- "Unknown"
current.all <- current.all %>% arrange(desc(Occupancy), WaterBody, Year, SiteName)

hist.all <- hist.all %>% 
  arrange(-hOccu, WaterBody) %>% 
  relocate(.,SiteName) %>% 
  relocate(.,WaterBody,.after = SiteName) %>% 
  relocate(.,hOccu,.after = WaterBody) %>% 
  relocate(.,Year,.after = hOccu) %>% 
  relocate(.,State,.after = Year) %>% 
  relocate(.,c(x, y),.after = State) %>% 
  relocate(.,PointType,.after = y) %>% 
  relocate(.,Source,.after = PointType) %>% 
  select(-Date, -WBType, -SiteCode, -SiteHist, -BS_Pres, -HStatus) %>% 
  rename(Easting = x, Northing = y) %>% 
  rename(Occupancy = hOccu)

hist.all[which(hist.all$Occupancy == -9999),]$Occupancy <- "Historical presence (likely misidentified)"
hist.all[which(hist.all$Occupancy == 1),]$Occupancy <- "Historical presence"
hist.all[which(hist.all$SiteName == "SEBAGO"),]$Occupancy <- "Historical presence"
hist.all[which(hist.all$Year == -9999),]$Year <- "Unknown"
hist.all <- hist.all %>% arrange(Occupancy, WaterBody, Year, SiteName)
```

Combine:
```{r}
all.sites <- rbind(hist.all, current.all)
```

Clean up sources for table:
```{r}
unique(all.sites$Source)

all.sites[which(all.sites$Source == "NH Fish StoryMap"),]$Source <- 1
all.sites[which(all.sites$Source == "M. Carpenter, NHFGD, written communication, 7 February 2023"),]$Source <- 2
all.sites[which(all.sites$Source == "Cooper 1939, M. Gallagher, MDIFW, written communication, 25 January 2021"),]$Source <- "3, 4"
all.sites[which(all.sites$Source == "M. Gallagher, MDIFW, written communication, 25 January 2021"),]$Source <- 4
all.sites[which(all.sites$Source == "NHFGD 2015 SWG Report"),]$Source <- 5
all.sites[which(all.sites$Source == "Kendall 1914"),]$Source <- 6
all.sites[which(all.sites$Source == "Fishmap.org"),]$Source <- 7
all.sites[which(all.sites$Source == "Doering et al. 1995"),]$Source <- 8
all.sites[which(all.sites$Source == "Harrington 1947b, NHFGD 2015 SWG Report"),]$Source <- "5, 9"
all.sites[which(all.sites$Source == "USEPA 2016"),]$Source <- 10
all.sites[which(all.sites$Source == "Harrington 1948"),]$Source <- 11
all.sites[which(all.sites$Source == "Maine Aquatic Biodiversity Project"),]$Source <- 12
all.sites[which(all.sites$Source == "University of Maine Bridle Shiner Surveys 2021-2022"),]$Source <- 13
all.sites[which(all.sites$Source == "Kendall 1914, M. Gallagher, MDIFW, written communication, 25 January 2021"),]$Source <- "4, 6"

sources <- c("Sources: (1) New Hampshire Fish and Game Department (NHFGD) n.d.; (2) M. Carpenter, NHFGD, written communication, 7 February 2023; (3) Cooper 1939; (4) M. Gallagher, Maine Department of Inland Fisheries and Wildlife (MDIFW), written communication, 25 January 2021; (5) NHFGD 2015; (6) Kendall 1914; (7) North American Native Fishes Association 2023; (8) Doering et al. 1995; (9) Harrington 1947b; (10) USEPA 2016, (11) Harrington 1948; (12) Vaux 2013; (13) University of Maine bridle shiner surveys 2021-2022 (this study).", rep(NA, ncol(all.sites)-1))

all.sites.sourced <- rbind(all.sites, sources)

write.csv(all.sites.sourced, "./Tables/SupplementD_TableSD1-update.csv")
```

# Old code, ignore

```{r}
# Function from https://github.com/josiahparry/sfdep/blob/HEAD/R/point-pattern-centers.R
# center_mean <- function(geometry, weights = NULL) {
#   geometry <- check_polygon(geometry)
#   crs <- sf::st_crs(geometry)
#   coords <- sf::st_coordinates(geometry)
# 
#   n <- nrow(coords)
# 
#   if (!is.null(weights)) {
#     # https://github.com/JosiahParry/sfdep/issues/35
#     # shouts out @JoseLastra
#     res <- colSums(coords * weights) / sum(weights, na.rm = TRUE)
#   } else {
#     res <- colSums(coords) / n
#   }
#   sf::st_sfc(sf::st_point(res), crs = crs)
# }
```
