---
title: "Download SoilGrid rasters"
author: "Lara Katz"
date: "2024-05-06"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# install.packages('XML')
# install.packages('rgdal')
# install.packages('gdalUtils')
# install.packages('sf')
# install.packages('dplyr')
# devtools::install_github("envirometrix/landmap")
# install.packages(leaflet)
# install.packages(mapview)
# library(devtools)
# devtools::install_github("https://github.com/be-marc/leaflet.opacity", dependencies=T)
library(XML)
library(rgdal)
library(gdalUtilities)
library(raster)
library(sf)
library(dplyr)
library(RColorBrewer)
library(leaflet)
library(leaflet.opacity)
library(mapview)
```

Make bounding box and get coordinates
```{r}
# Bounding box 
bbcoordsst <- st_read("./datos/BoundingBoxHomolosine.shp")
igh='+proj=igh +lat_0=0 +lon_0=0 +datum=WGS84 +units=m +no_defs' # proj string for Homolosine projection
ne_igh <- st_transform(bbcoordsst, igh)

(bbox <- st_bbox(ne_igh))
# xmin     ymin     xmax     ymax 
#-9137229  4520021 -8362878  5390855
```

Use the bbox data to define the boundary box limits as used by GDAL
```{r}
## ul means upper left
## lr means lower right
ulx = bbox$xmin
uly = bbox$ymax
lrx = bbox$xmax
lry = bbox$ymin

(bb <- c(ulx, uly, lrx, lry))
#     xmin     ymax     xmax     ymin 
# -9137229  5390855 -8362878  4520021 
```
SoilGrids maps have associated uncertainties as any product derived from a modelling approach. The prediction uncertainty is quantified by probability distributions. For each property and each standard depth interval this distribution is characterised by four parameters as follows:
Q0.05 - 5% quantile;
Q0.50 - median of the distribution;
mean - mean of the distribution;
Q0.95 - 95% quantile.

## Clay content raster:

Define variable of interest, depth, and quantile:
```{r}
voi = "clay" # variable of interest
depth = "0-5cm"
quantile = "mean"

(layer = paste(voi, paste(voi,depth,quantile, sep="_"), sep = "/")) # layer of interest 

(vrt_layer = paste(layer, '.vrt', sep=""))
```

Use the gdal_translate function to download the clay layer.
```{r}
sg_url="/vsicurl/https://files.isric.org/soilgrids/latest/data/"
lfile = "./datos/clay_ntest_igh_0_5.vrt"
```

Obtain a VRT for the area of interest in the Homolosine projection. We suggest to use VRT for the intermediate steps to save space and computation times.
```{r}
gdal_translate(paste0(sg_url,vrt_layer), lfile ,
               of = "VRT",
               tr=c(250,250), # data is 250m x 250m
               projwin=bb,
               projwin_srs =igh)
```

Change projection to NAD83 / 19N
```{r}
gdalwarp(lfile,
         "./datos/clay_19N.vrt", 
         s_srs=igh, 
         t_srs="EPSG:26919", 
         of="VRT",
         overwrite = TRUE)
```

Make final Geotiff
```{r}
gdal_translate("./datos/clay_19N.vrt",  
               "./datos/clay_19N.tif", 
               co=c("TILED=YES","COMPRESS=DEFLATE","PREDICTOR=2","BIGTIFF=YES"))
```

## Silt raster

Define variable of interest, depth, and quantile 
```{r}
voi = "silt" # variable of interest
depth = "0-5cm"
quantile = "mean"

(layer = paste(voi, paste(voi,depth,quantile, sep="_"), sep = "/")) # layer of interest 

(vrt_layer = paste(layer, '.vrt', sep=""))
```

Use the gdal_translate function to download the silt layer.
```{r}
sg_url="/vsicurl/https://files.isric.org/soilgrids/latest/data/"
lfile = "./datos/silt_igh_0_5.vrt"
```

Obtain a VRT for the area of interest in the Homolosine projection. Use VRT for the intermediate steps to save space and computation times.
```{r}
gdal_translate(paste0(sg_url,vrt_layer), lfile ,
               of = "VRT",
               tr=c(250,250), # data is 250m x 250m
               projwin=bb,
               projwin_srs =igh)
```

Change projection to NAD83 / 19N
```{r}
gdalwarp(lfile,
         "./datos/silt_19N.vrt", 
         s_srs=igh, 
         t_srs="EPSG:26919", 
         of="VRT",
         overwrite = TRUE)
```

Make final Geotiff
```{r}
gdal_translate("./datos/silt_19N.vrt",  
               "./datos/SiltContent_19N.tif", 
               co=c("TILED=YES","COMPRESS=DEFLATE","PREDICTOR=2","BIGTIFF=YES"))
```

# Sand raster

Define variable of interest, depth, and quantile 
```{r}
voi = "sand" # variable of interest
depth = "0-5cm"
quantile = "mean"

(layer = paste(voi, paste(voi,depth,quantile, sep="_"), sep = "/")) # layer of interest 

(vrt_layer = paste(layer, '.vrt', sep=""))
```

Use the gdal_translate function to download the sand layer.
```{r}
sg_url="/vsicurl/https://files.isric.org/soilgrids/latest/data/"
lfile = "./datos/sand_igh_0_5.vrt"
```

Obtain a VRT for the area of interest in the Homolosine projection. Use VRT for the intermediate steps to save space and computation times.
```{r}
gdal_translate(paste0(sg_url,vrt_layer), lfile,
               of = "VRT",
               tr=c(250,250), # data is 250m x 250m
               projwin=bb,
               projwin_srs =igh)
```

Change projection to NAD83 / 19N
```{r}
gdalwarp(lfile,
         "./datos/sand_19N.vrt", 
         s_srs=igh, 
         t_srs="EPSG:26919", 
         of="VRT",
         overwrite = TRUE)
```

Make final Geotiff
```{r}
gdal_translate("./datos/sand_19N.vrt",  
               "./datos/SandContent_19N.tif", 
               co=c("TILED=YES","COMPRESS=DEFLATE","PREDICTOR=2","BIGTIFF=YES"))
```

# Coarse fragments raster

Define variable of interest, depth, and quantile:
```{r}
voi = "cfvo" # variable of interest
depth = "0-5cm"
quantile = "mean"

(layer = paste(voi, paste(voi,depth,quantile, sep="_"), sep = "/")) # layer of interest 

(vrt_layer = paste(layer, '.vrt', sep=""))
```

Use the gdal_translate function to download the coarse fragments layer.
```{r}
sg_url="/vsicurl/https://files.isric.org/soilgrids/latest/data/"
lfile = "./datos/cfvo_igh_0_5.vrt"
```

Obtain a VRT for the area of interest in the Homolosine projection. Use VRT for the intermediate steps to save space and computation times.
```{r}
gdal_translate(paste0(sg_url,vrt_layer), lfile ,
               of = "VRT",
               tr=c(250,250), # data is 250m x 250m
               projwin=bb,
               projwin_srs =igh)
```

Change projection to NAD83 / 19N
```{r}
gdalwarp(lfile,
         "./datos/CoarseFrag_19N.vrt", 
         s_srs=igh, 
         t_srs="EPSG:26919", 
         of="VRT",
         overwrite = TRUE)
```

Make final Geotiff:
```{r}
gdal_translate("./datos/CoarseFrag_19N.vrt",  
               "./datos/CoarseFrag_19N.tif", 
               co=c("TILED=YES","COMPRESS=DEFLATE","PREDICTOR=2","BIGTIFF=YES"))

```

